---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r, echo=TRUE} 
unzip("activity.zip")
initialData <- read.csv("activity.csv", colClasses=c("numeric", "Date", "numeric"))
```

```{r, echo=TRUE} 
head(initialData)
str(initialData)
```


## What is mean total number of steps taken per day?
``{r, echo=TRUE}
data <- initialData[!(is.na(initialData$steps)), ]
```

Calculate total no of steps:

```{r, echo=TRUE}
totalStepsDay <- aggregate(steps ~ date, data, sum)
head(totalStepsDay)
```

Creating plots

```{r, echo=TRUE}
paletteBlue <- colorRampPalette(c("skyblue", "darkblue", "skyblue"))
hist(totalStepsDay$steps, breaks=20, xlab="Number of Steps Taken", 
     main="Histogram of the Total Number of Steps Taken per Day",
     col=paletteBlue(22), family="serif")
```

```{r, message=FALSE}
library(dplyr)
totalStepsSummary <- summarise(totalStepsDay, meanOfTotalSteps=mean(totalStepsDay$steps),
                               medianOfTotalSteps=median(totalStepsDay$steps))
print(totalStepsSummary)
```

mean value calculated is **10766.19**, and the median value **10765**.

## What is the average daily activity pattern?

```{r, echo=TRUE}
meanStepsInterval <- aggregate(steps ~ interval, data, mean)
head(meanStepsInterval)
```


```{r, echo=TRUE}
plot(x=meanStepsInterval$interval, y=meanStepsInterval$steps, type="l",
     main="Time Series Plot of Average Steps Taken per Interval",
     ylab="Number of Steps", xlab="Intervals (in 5 mins)",
     col="darkblue", lwd=1.5, family="serif")
```


```{r, echo=TRUE}
meanStepsInterval[grep(max(meanStepsInterval$steps), meanStepsInterval$steps), ]
```
interval with the maximum number of steps is interval **835**.

## Imputing missing values
```{r, echo=TRUE}
anyNA(initialData)
```

```{r, echo=TRUE}
data.frame(steps=sum(is.na(initialData$steps)), 
           interval=sum(is.na(initialData$interval)), 
           date=sum(is.na(initialData$date)))
```

```{r, echo=TRUE}
imputedData <- initialData
for(x in 1:17568) {
    if(is.na(imputedData[x, 1])==TRUE) {
        imputedData[x, 1] <- meanStepsInterval[meanStepsInterval$interval %in% imputedData[x, 3], 2]
    }
}
head(imputedData)
```

```{r, echo=TRUE}
imputedTotalStepsDay <- aggregate(steps ~ date, imputedData, sum)
head(imputedTotalStepsDay)
```

Creating the histogram:

```{r, echo=TRUE}
paletteRed <- colorRampPalette(c("deeppink", "darkred", "deeppink"))
hist(imputedTotalStepsDay$steps, breaks=20, xlab="Number of Steps Taken", 
     main="Histogram of Total Number of Steps Taken per Day (With Imputed Values)",
     col=paletteRed(22), family="serif")
```

```{r, echo=TRUE}
imputedStepsSummary <- summarise(imputedTotalStepsDay, 
                                 meanOfTotalSteps=mean(imputedTotalStepsDay$steps), 
                                 medianOfTotalSteps=median(imputedTotalStepsDay$steps))  
print(imputedStepsSummary)
```

```{r, echo=TRUE}
rbind(totalStepsSummary, imputedTotalStepsSummary)
```

If histograms of the two data sets (imputed and non-imputed) are compared:

```{r, echo=TRUE}
par(mfrow = c(1, 2))
hist(totalStepsDay$steps, breaks=20, xlab="Number of Steps Taken", 
     col=paletteBlue(22), family="serif", ylim=c(0, 20), main=NULL)
hist(imputedTotalStepsDay$steps, breaks=20, xlab="Number of Steps Taken", 
     col=paletteRed(22), family="serif", ylim=c(0, 20), main=NULL)
mtext("Histograms of Total Number of Steps Taken per Day, Without/With Imputed Values",
      adj=0.95, family="serif", font=2)
```

```{r, echo=TRUE}
naByDate <- initialData[is.na(initialData$steps), ]
table(naByDate$date)
```

```{r, echo=TRUE}
length(unique(data$interval))
```



## Are there differences in activity patterns between weekdays and weekends?
```{r, echo=TRUE}
daysData <- imputedData
daysData$days <- weekdays(daysData$date)
daysData$weekday <- as.character(rep(0, times=17568))
for(x in 1:17568) {
    if(daysData[x, 4] %in% c("Saturday", "Sunday")) {
        daysData[x, 5] <- "weekend"
    } else {
        daysData[x, 5] <- "weekday"
    }
}
daysData$weekday <- factor(daysData$weekday)
head(daysData)
```

To compare the weekday and weekend data, and create two plots of the average number of steps taken per interval, the data has to be split into two groups of weekday/weekend data, using the newly created variable.

```{r, echo=TRUE}
weekdayData <- daysData[daysData$weekday=="weekday", ]
weekendData <- daysData[daysData$weekday=="weekend", ]
```

```{r, echo=TRUE}
weekdayMean <- aggregate(steps ~ interval, weekdayData, mean)
weekendMean <- aggregate(steps ~ interval, weekendData, mean)
```

plot:
```{r, echo=TRUE}
par(mfrow=c(2, 1), mar=c(4, 4.1, 3, 2.1))
plot(weekdayMean$interval, weekdayMean$steps, type="l",
     main="Time Series Plot of Average Steps Taken per Interval, for Weekdays",
     xlab="Intervals (in 5 mins)", ylab="Number of Steps", family="serif",
     col="darkred", lwd=1.5, ylim=c(0, 230))
plot(weekendMean$interval, weekendMean$steps, type="l",
     main="Time Series Plot of Average Steps Taken per Interval, for Weekends",
     xlab="Intervals (in 5 mins)", ylab="Number of Steps", family="serif",
     col="darkblue", lwd=1.5, ylim=c(0, 230))
```
